---
title: "Competing Risks Survival Forests with ranger"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Competing Risks Survival Forests with ranger}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

Ranger supports competing risks survival forests, where subjects can experience one of several mutually exclusive event types.
The implementation follows the cause-specific hazard approach described in Ishwaran et al. (2014), similar to the `randomForestSRC` package.

In a competing risks setting, the status variable takes values:

- `0` = censored (no event observed)
- `1, 2, ..., K` = event of type 1, 2, ..., K

For example, in a study of cancer patients, death from cancer (status = 1) and death from other causes (status = 2) are competing events.

## Method

**Splitting:** At each node, ranger computes a cause-specific log-rank statistic for each event type, treating all other events as censored.
These are combined into a composite statistic: $\sqrt{\sum_{j=1}^K w_j Z_j^2}$, where $Z_j$ is the log-rank Z-statistic for event type $j$ and $w_j$ are equal cause weights.
This composite test finds splits that are informative for discriminating risk across all event types simultaneously.

**Predictions:** Each terminal node stores a cause-specific cumulative hazard function (CHF) per event type via the Nelson-Aalen estimator.
The cumulative incidence functions (CIF) are then computed from the cause-specific CHFs using the Aalen-Johansen estimator:
$$\text{CIF}_j(t) = \sum_{s \le t} \hat{S}(s^{-}) \cdot \hat{h}_j(s)$$
where $\hat{S}(t^{-})$ is the event-free survival probability just before time $t$, and $\hat{h}_j(s)$ is the cause-specific hazard increment for event $j$.

## Quick start

```{r basic}
library(ranger)
library(survival)

## Simulate competing risks data
set.seed(42)
n <- 500
x1 <- rnorm(n)
x2 <- rnorm(n)

## Latent event times for two competing causes
t1 <- rexp(n, rate = exp(0.5 * x1))   # cause 1 depends on x1
t2 <- rexp(n, rate = exp(0.3 * x2))   # cause 2 depends on x2
tc <- rexp(n, rate = 0.3)             # censoring

time   <- pmin(t1, t2, tc)
status <- ifelse(time == tc, 0L, ifelse(time == t1, 1L, 2L))

dat <- data.frame(time = time, status = status, x1 = x1, x2 = x2)
table(dat$status)
```

Fit a competing risks forest using the standard `Surv()` formula interface:

```{r fit}
rf <- ranger(Surv(time, status) ~ ., data = dat, num.trees = 500)
rf
```

The `num.event.types` field confirms competing risks were detected:

```{r check}
rf$num.event.types
```

## Output structure

For competing risks, `chf` and `cif` are lists of matrices (one per event type), each of dimension (n_samples x n_timepoints):

```{r output}
## Cause-specific CHFs
length(rf$chf)
dim(rf$chf[[1]])

## Cumulative incidence functions
length(rf$cif)
dim(rf$cif[[1]])
```

Access the CIF for event type 1 for the first sample:

```{r access}
head(rf$cif[[1]][1, ])
```

## Plotting CIF curves

```{r plot_mean, fig.cap = "Mean CIF across all subjects for both event types."}
times <- rf$unique.death.times

par(mar = c(4.5, 4.5, 2, 1))
plot(times, colMeans(rf$cif[[1]]), type = "s", col = "steelblue", lwd = 2,
     xlab = "Time", ylab = "Cumulative Incidence", ylim = c(0, 0.7))
lines(times, colMeans(rf$cif[[2]]), type = "s", col = "firebrick", lwd = 2)
legend("bottomright", legend = c("Event 1", "Event 2"),
       col = c("steelblue", "firebrick"), lwd = 2, bty = "n")
```

Individual subject CIF curves for a high-risk and low-risk subject:

```{r plot_individual, fig.cap = "CIF curves for individual subjects."}
i_high <- which.max(dat$x1)
i_low  <- which.min(dat$x1)

par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

plot(times, rf$cif[[1]][i_high, ], type = "s", col = "steelblue", lwd = 2,
     xlab = "Time", ylab = "CIF (Event 1)",
     main = sprintf("High risk (x1 = %.1f)", dat$x1[i_high]), ylim = c(0, 1))

plot(times, rf$cif[[1]][i_low, ], type = "s", col = "steelblue", lwd = 2,
     xlab = "Time", ylab = "CIF (Event 1)",
     main = sprintf("Low risk (x1 = %.1f)", dat$x1[i_low]), ylim = c(0, 1))
```

## Prediction on new data

```{r predict}
newdat <- data.frame(x1 = c(-1, 0, 1), x2 = c(0, 0, 0))
pred <- predict(rf, data = newdat)

## Predictions have the same structure
length(pred$cif)
dim(pred$cif[[1]])
```

```{r plot_pred, fig.cap = "Predicted CIF for event 1 for three covariate profiles."}
par(mar = c(4.5, 4.5, 2, 1))
cols <- c("forestgreen", "orange", "firebrick")
plot(NULL, xlim = range(pred$unique.death.times), ylim = c(0, 1),
     xlab = "Time", ylab = "CIF (Event 1)")
for (i in 1:3) {
  lines(pred$unique.death.times, pred$cif[[1]][i, ],
        type = "s", col = cols[i], lwd = 2)
}
legend("bottomright", legend = paste0("x1 = ", c(-1, 0, 1)),
       col = cols, lwd = 2, bty = "n")
```

## Alternative interface

The `dependent.variable.name` / `status.variable.name` interface also works:

```{r alternative}
rf2 <- ranger(dependent.variable.name = "time",
              status.variable.name = "status",
              data = dat, num.trees = 100)
rf2$num.event.types
```

## Variable importance

Variable importance works as usual:

```{r importance}
rf_imp <- ranger(Surv(time, status) ~ ., data = dat,
                 num.trees = 500, importance = "permutation")
rf_imp$variable.importance
```

## Split rules

The following split rules are supported for competing risks:

- `"logrank"` (default) -- composite cause-specific log-rank
- `"extratrees"` -- extremely randomized trees with composite log-rank
- `"maxstat"` -- maximally selected rank statistics

The C-index (`"C"`) split rule is not supported for competing risks.

```{r splitrules}
rf_et <- ranger(Surv(time, status) ~ ., data = dat,
                num.trees = 100, splitrule = "extratrees")
rf_et$num.event.types
```

## Backward compatibility

For standard survival data (binary status 0/1), ranger behaves exactly as before -- `chf` and `survival` are matrices and `cif` is `NULL`:

```{r compat}
dat_std <- dat
dat_std$status <- ifelse(dat$status > 0, 1, 0)

rf_std <- ranger(Surv(time, status) ~ ., data = dat_std, num.trees = 100)
is.matrix(rf_std$chf)
is.matrix(rf_std$survival)
is.null(rf_std$cif)
```

## References

- Ishwaran, H., Gerds, T. A., Kogalur, U. B., Moore, R. D., Gange, S. J. & Lau, B. M. (2014). Random survival forests for competing risks. *Biostatistics*, 15(4):757-773. doi:10.1093/biostatistics/kxu010.
- Ishwaran, H., Kogalur, U. B., Blackstone, E. H. & Lauer, M. S. (2008). Random survival forests. *Annals of Applied Statistics*, 2(3):841-860.
- Wright, M. N. & Ziegler, A. (2017). ranger: A fast implementation of random forests for high dimensional data in C++ and R. *Journal of Statistical Software*, 77(1):1-17. doi:10.18637/jss.v077.i01.
